RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-12-31 manuscripts/manuscript.Rmd
First part of metadata section rewrite.

The data components being added to a `nexml` object need not share the
same OTUs.  For example, here we append additional phylogenies and characters
using different OTUs to the `nexml` object originally populated with
the `geospiza` phylogeny and character data:

```{r}
data(primates)
nexObj <- add_trees(primates$phy, nexObj)
nexObj <- add_characters(primates$dat, nexObj)
```

`RNeXML` automatically adds new, separate OTU blocks into the NeXML file
for each of these new data sets.
Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML file. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies or
character states in one NeXML file.

# Metadata

NeXML allows attaching metadata to any data element, and even to
metadata themselves. At the level of the document as a whole or the
phylogeny, metadata can provide bibliographic and provenance
information, for example about the study as part of which the
phylogeny was generated or applied, and with which methods it was
generated. In the form of annotations, metadata can also be attached
to very specific elements of the data, such as individual OTUs, nodes,
or even edges of the phylogeny.

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations" (RDFa),
as described in @Vos_2012 or @W3C_2014. 
These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

## Writing NeXML metadata

The burden typically involved with documenting and sharing good
metadata alongside the data remains a
significant barrier to increased sharing and reuse of scientific research
[@Tenopir_2011]. To promote sharing of well-documented phylogenetic data,
`RNeXML` aims to provide a simple interface to adding well-defined machine-readable metadata to NeXML files. This occurs on
several levels:

- Automatic annotation
- Adding common metadata fields (`add_basic_metadata`)
- Taxonomic identifiers (`taxize_nexml`)
- Generic metadata (`add_meta`)
- Extending the NeXML standard using RDFa

`RNeXML`'s `nexml_write` function adds certain basic metadata by automatically
if it has not yet been provided; using default values
consistent with recommended best practices [@Cranston_2014]. At this time,
this includes naming the software generating the NeXML, a time-stamp of
when a tree was produced, and an open data license.

To adjust these values and also add other metadata fields commonly recommended for
includion with publicly archived data, `RNeXML` provides the `add_basic_meta()`
function.  This function hides the complexities of anchoring metadata
attribute names in common vocabularies from the user.  The
`add_basic_meta()` function takes as input an existing `nexml` object
(like the other `add_` functions, if none is provided it will create one), and at the time
of this writing any of the following
parameters: `title`, `description`, `creator`, `pubdate`, `rights`, `publisher`,
`citation`.  Other metadata elements and corresponding parameters may
be added in the future.

Here we create a `nexml` object for the `bird.orders` data
provided in the `ape` package, and then add basic metadata:

```{r}
data(bird.orders)
birds <- add_trees(bird.orders)
birds <- add_basic_meta(
  title = "Phylogeny of the Orders of Birds From Sibley and Ahlquist",

  description = "This data set describes the phylogenetic relationships of the
     orders of birds as reported by Sibley and Ahlquist (1990). Sibley
     and Ahlquist inferred this phylogeny from an extensive number of
     DNA/DNA hybridization experiments. The ``tapestry'' reported by
     these two authors (more than 1000 species out of the ca. 9000
     extant bird species) generated a lot of debates.

     The present tree is based on the relationships among orders. The
     branch lengths were calculated from the values of Delta T50H as
     found in Sibley and Ahlquist (1990, fig. 353).",

  citation = "Sibley, C. G. and Ahlquist, J. E. (1990) Phylogeny and
     classification of birds: a study in molecular evolution. New
     Haven: Yale University Press.",

  creator = "Sibley, C. G. and Ahlquist, J. E.",
  nexml = birds)
```

<!-- the above will get fixed by CB, this is just a reminder and -->
<!-- placeholder to adjust when that's done. -->
<!-- Um, sorry, remind me what I am supposed to fix here? formatting? -->

Instead of a literal string, citations can also be provided in R's
`bibentry` type, which is the one in which R package citations are obtained:

```{r}
birds <- add_basic_meta(citation = citation("ape"), nexml=birds)
```

Alternatively, a citation to a published paper with a Digital
Object Identifier (DOI) can be provided in the form of the DOI,
which the package `knitcitations` [@Boettiger_2014] can turn into a
formatted citation using. As an example, to add the above citation
using the DOI:

```{r }
library("knitcitations")
birds <- add_basic_meta(citation = bib_metadata("10.1093/bioinformatics/btg412"), nexml=birds)
```

## Taxonomic identifiers

`RNeXML` provides various functions to assist in automatically adding
other more specific metadata to various elements of the NeXML data.
One common use case is that of taxonomic identifiers.  Misspellings or
inconsistent use of taxonomic names in phylogenetic data are a common
source of frustration when reusing phylogenetic trees, especially when
integrating data from different sources: trees might come from one place
that uses one taxonomy, character data and occurrence data might come from
other places that use different taxonomies. Hence, not just misspellings,
but also synonyms, homonyms and incompatible taxon concepts might plague
the end user. By associating the taxonomic labels against a specific,
centralized database we can catch some of these inconsistencies and
indicate more precisely which classification we are appealing to in the
event of conflicting taxonomic definitions.  In this example, we indicate
that we intend the bird orders listed in the tree to correspond to the
NCBI definitions.

The `taxize_nexml()` function uses the R package `taxize` 
[@Chamberlain_2013] to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.

```{r message=FALSE, results='hide'}
birds <- taxize_nexml(birds, "NCBI")
```
If no match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided. If multiple matches are found,
the user will be prompted to choose between them.

## Custom metadata extensions

So far our examples have relied on functions that have a built-in
understanding of the kind of metadata being provided, and can thus
generate the appropriate metadata with little further guidance.
In order to be machine-readable and understandable, NeXML requires
all metadata annotations follow the RDFa format (see @Vos_2012) which
explicitly declares the meaning of the annotation.

Some basic definitions about the XML syntax used to express NeXML
will be helpful for this discussion, though a more comprehensive 
introduction can be found in @Vos_2012.  For instance, consider how
the title we gave to the `bird.orders` phylogeny appears in the NeXML file
using the XML format:

```xml
<meta id="m2" property="dc:title" datatype="xsd:string"
      content="Phylogeny of the Orders of Birds From Sibley and Ahlquist"/> 
```

The format begins with an XML "tag" and the name "meta", indicated by the opening
angle bracket; `<meta`.  A closing tag `\>` indicates the end of this
XML "element" .  This tag contains several "attributes", named "id",
"property", etc., that an R user can think of as names in a `list`. What
follows the equals sign after the attribute name is called the attribute "value",
(e.g. the `id` attribute has value "m2").

The name of any element (like "meta") or attribute (like "id") in XML must be associated
with a particular dictionary called a "namespace" and specified using a "prefix"
of an abbreviation followed by a colon appearing before the name.  If no prefix
is given, as is the case for all attributes here: `id`, `property`, `datatype`,
and `content`, then a default prefix specified at the beginning of the document applies,
in this case, the NeXML namespace.  This allows users to combine different XML
that have different definitions of terms like `id` or `property`.

The example element above also uses namespaces on some of the attribute values themselves.
This is a consequence of RDFa annotation, and not strictly part of XML, though
it works much the same way.  While we hope the following examples will make these
concepts clearer, users do not need to understand XML and namespaces to use most
of the basic annotation features provided by `RNeXML`.

So much for the structure; we can now briefly explain the meaning of this element. 
We see the title appears as the value of the `content` attribute of this `meta` element.
The `id` attribute provides us with a unique way to refer to this meta
element in this document.  The `datatype` tells us that the content
is a text string (as opposed to, say, a numerical or boolean value).
The `property` is the attribute that most interests us here; it tells us
that the content is a title.  The prefix `dc` indicates the vocabulary
(here identified as an XML namespace) to which the word `title` belongs.  The prefix
is defined elsewhere in the NeXML file, as we can see with:

>>>>

The data with which a `nexml` object is populated need not share the
same OTUs.  `RNeXML` automatically adds new, separate OTU blocks into
the NeXML file for each data matrix and tree that uses a set of OTUs
that differs from others.

Other than storage size, there is no limit to the number of
phylogenies and character matrices that can be included in a
single NeXML file. This allows, for example, to capture samples from a
posterior probability distribution of inferred or simulated phylogenies or
character states in one NeXML file.

## Data documentation and annotation with built-in metadata

NeXML allows attaching metadata ("_annotations_") to any data element,
and even to metadata themselves. Whether at the level of the document
as a whole or the an individual data matrix or phylogeny, metadata can
provide bibliographic and provenance information, for example about
the study as part of which the phylogeny was generated or applied,
which data matrix and which methods were used to generate it. Metadata
can also be attached to very specific elements of the data, such as
specific traits, individual OTUs, nodes, or even edges of the
phylogeny.

NeXML uses the "Resource Description Framework in Annotations" (RDFa),
as described in @Vos_2012 or @W3C_2014, to encode metadata
annotations. This standard provides a precisely specified,
machine-readable format that maintains backwards compatibility with
compliant NeXML parsers (and thus `RNeXML`), because the capacity of a
tool to _parse_ annotations is not predicated on _understanding_
annotations it has not seen before.

The burden typically involved with documenting and sharing good
metadata alongside the data remains a significant barrier to increased
sharing and reuse of scientific research [@Tenopir_2011]. To promote
sharing of well-documented phylogenetic data, `RNeXML` aims to
alleviate the burden to recording useful and machine-readable metadata
at several levels.

- Automatic annotation: 
- Adding common metadata fields (`add_basic_metadata`)
- Taxonomic identifiers (`taxize_nexml`)
- Generic metadata (`add_meta`)
- Extending the NeXML standard using RDFa

First, when writing a NeXML file the package adds certain basic
metadata automatically if it has not yet been provided, using default
values consistent with recommended best practices
[@Cranston_2014]. Currently, this includes naming the software
generating the NeXML, a time-stamp of when a tree was produced, and an
open data license.

Second, `RNeXML` provides a simple method, called
`add_basic_metadata()`, to set metadata attributes commonly
recommended to be included with publicly archived or shared data
[@Cranston_2014]. The currently accepted parameters include `title`,
`description`, `creator`, `pubdate`, `rights`, `publisher`, and
`citation`. Behind the scenes the method automatically anchors these
attributes in common vocabularies (such as Dublin Core).

Third, `RNeXML` includes methods to assist with common cases of adding
metadata to specific elements of the NeXML data that increase the
data's utility for reuse. For example, misspellings and inconsistent
taxonomic names are a frequent source of frustration especially when
recombining phylogenetic data from different sources. These obstacles
can be mitigated by annotating the taxonomic names with identifiers,
which also allows being explicit about the identifier authority being
used. `RNeXML` includes the `taxize_nexml()` method, which uses the R
package `taxize` [@Chamberlain_2013] to match each taxon label against
the NCBI database, and, provided a unique match is found, annotates
the respective OTU with the matching NCBI identifier,.

## Data annotation with custom metadata extensions

The RDFa standard adopted by NeXML (see @Vos_2012) for all metadata
annotations defines the structure and syntax for such annotations, but
does not restrict the content. Specifically, every metadata annotation
consists of a metadata property and a value of the property; the
bearer of the property is determined by where the annotation appears
in the NeXML document (technically, the containing element in the XML
document model). Technically, this structure corresponds directly to
Resource Description Framework (RDF) subject-predicate-object triples.
It therefore allows for arbitrary custom extensions to meet ongoing
scientific needs without requiring modifications (or "extensions") of
the syntax, provided the metadata annotations are cast as RDF triples.

The `RNeXML` methods described above for built-in metadata annotations
perform this cast automatically behind the scenes, including mapping
metadata attributes to terms in requisite common vocabularies (such as
Dublin Core for "title", "creator", etc.) or ontologies. Giving
metadata properties (and sometimes values) as vocabulary or ontology
terms rather than simple text strings is crucial for allowing machines
to not only parse but also interpret their semantics. To achieve this
benefit for custom metadata extensions, the user necessarily needs to
handle certain technical details from which the `RNeXML` API shields
her otherwise, in particular the vocabularies or ontologies of
metadata terms in the form of globally unique identifiers (normally
HTTP URIs). To be consistent with XML terminology for more technical
users, `RNeXML` calls these _namespaces_, and their abbreviations
_prefixes_. For example, the namespace for the Dublin Core Metadata
Terms vocabulary, commonly abbreviated with "dc", is
"http://purl.org/dc/elements/1.1/". Using its abbreviation, a metadata
property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title", the identifier for the term
"title" in said vocabulary.

`RNeXML` includes methods to obtain and manipulate these values. The
`get_namespaces()` method accepts a `nexml` object and returns a list
with element names being the namespace abbreviations and element
values being the corresponding identifies (typically HTTP URIs):
